Describe Participants: Characteristics, Diet, and Cardiometabolic Risk

### 2022-01-05 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Add dietary inflammation index score

@@

Is TMAO significantly different by ethnicity?
```{r Ethnicity}
ethnicity_aov <- (aov(tmao_log ~ ethnicity, phen))
summary(ethnicity_aov)
TukeyHSD(ethnicity_aov)

@@

There are significant differences in TMAO levels by ethnic groups but ethnicity does not predict TMAO.

Back to tables... make similar tables but show TMAO quantiles

```{r Quantile TMAO}
# Quantile
tmao_quantile <- quantile(phen$tmao)
phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))

@@

@@

Save table
```{r Save TMAO linear regression table}
# Save table
write.csv(res, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD.csv")
```

## Females 
```{r Cardiometabolic Multiple Linear Regression - Females}
# Transfored variable list
transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln",  "insulin_bd1_ln", "tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")

@@

Save table
```{r Save TMAO linear regression table - Females}
# Save table
write.csv(res.f, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_females.csv")
```

@@

## Males 
```{r Cardiometabolic Multiple Linear Regression - Males}
# Transfored variable list
transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln", "insulin_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")

@@

@@

Save
```{r save food linear regression table}
# Save table
write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24.csv")
```

Make TMAO food intake by quantile table
```{r Food Intake by TMAO quantile}
# TMAO table by age bin - females
factorVars <- c("age_cat", "tmao_quantile")
vars <- c("pf_meat_totals", 
          "pf_curedmeat_totals",
          "pf_poult_totals",
          "pf_seafd_hi_totals",
          "pf_seafd_low_totals",
          "pf_eggs_totals",
          "pf_nutsds_totals",
          "pf_legumes_totals",
          "pf_total_totals",
          #"pf_soy_totals_Binary",
          "v_drkgr_totals",
          "v_redor_total_totals",
          #"v_redor_tomato_totals",
          #"v_redor_other_totals",
          "v_starchy_total_totals",
          #"v_starchy_potato_totals",
          #"v_starchy_other_totals",
          "v_other_totals",
          "v_total_totals",
          #"v_legumes_total",
          "g_whole_totals",
          "g_refined_totals",
          "g_total_totals",
          "d_total_totals",
          "d_milk_totals",
          "d_cheese_totals",
          "d_yogurt_totals",
          "hei_asa24_totalscore",
          "hei_asa24_totalveg",
          "hei_asa24_sfat",
          "choln_tnfs",
          "age",
          "ethnicity" ,
          "bmi_final",
          "tmao",
          "choline",
          "betaine",
          "carnitine")
table_food.fq <- CreateTableOne(vars = vars,
                                    strata = "tmao_quantile",
                                    data = full[full$sex == "Female",],
                                    factorVars = factorVars)
table_food.fq

>>>>

Create a table of participant characteristics by age category in males, females, and by sex.

@@

Is TMAO significantly different by ethnicity?

```{r Ethnicity}
ethnicity_aov <- (aov(tmao_log ~ ethnicity, phen))
summary(ethnicity_aov)
TukeyHSD(ethnicity_aov)

@@

There are significant differences in TMAO levels by ethnic groups but ethnicity does not predict TMAO when you control for the covariates sex*age, and cystatin C.

Create categorical TMAO groups by quantile and tertile. These will be useful when we are making tables and conducting categorical analyses related to TMAO.

Also, transform select variables so that they comply to the normal distribution (they have been tested by the Shapiro Wilk's test in previous analyses).

```{r Quantile and Tertile TMAO}
# Quantile
tmao_quantile <- quantile(phen$tmao)
phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))

@@

TMAO Quantile Tables

@@

Save table

```{r Save TMAO linear regression table}
# Save table
write.csv(res, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD.csv")
```

## Females

```{r Cardiometabolic Multiple Linear Regression - Females}
# Transfored variable list
transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln",  "insulin_bd1_ln", "tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")

@@

Save table

```{r Save TMAO linear regression table - Females}
# Save table
write.csv(res.f, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_females.csv")
```

@@

## Males

```{r Cardiometabolic Multiple Linear Regression - Males}
# Transfored variable list
transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln", "insulin_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")

@@

Load the Dietary Inflammation Index Score

```{r Dietary Inflammation Index}
# Load the diet scores from Levi and grep for the dietary inflammation index scores only
inflm <- read.csv("../../data/raw/from_LeviEvans/HEI scores_for LE_6142021.csv", row.names = 1)
inflm <- inflm[,grep("dii", colnames(inflm))]

# Combine with the full data from above
full <- merge(full, inflm, by = 0)
row.names(full) <- full[,"Row.names"]
full <- subset(full, select = -c(Row.names))
```

@@

```{r save food linear regression table}
# Save table
write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24.csv")
```

Check the relationship between the dietary inflammation index and fasting plasma TMAO.

```{r TMAO ~ dii - LCMS BD1 - lm 356 ppl, results='hide'}
transformed_keep_tbl2 <- c("sex", 
#"sex_numeric",
"kcal_tnfs", 
"age", 
"cystatinc_bd1", 
"ethnicity",
"tmao", 
"tmao_log", 
"dii_alcohol",
"dii_b12",
"dii_b6",
"dii_betacarotene",
"dii_caffeine",
"dii_carb",
"dii_chol",
"dii_kcal",
"dii_totalfat",
"dii_fiber",
"dii_folicacid",
"dii_iron",
"dii_magnesium",
"dii_mufa",
"dii_niacin",
"dii_omega3",
"dii_omega6",
"dii_protein",
"dii_pufa",
"dii_riboflavin",
"dii_satfat",
"dii_selenium",
"dii_thiamin",
"dii_transfat",
"dii_vita",
"dii_vitc",
"dii_vitd",
"dii_vite",
"dii_zinc",
"dii_tea" ,
"dii_flav3ol",
"dii_flavone",
"dii_flavonol",
"dii_flavonon",
"dii_anthocy",     
"dii_isoflav",
"dii_total_score")

# Make a smaller df from master (the old data)
full_sub <- full[, colnames(full) %in% transformed_keep_tbl2]
dii <- subset(full_sub, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
n <- ncol(dii)

# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii[,x] + sex*age + cystatinc_bd1 + kcal_tnfs, full_sub))

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm356_lcms) <- colnames(dii)
res_dii <- as.data.frame(my_lm356_lcms)

# BH correction
res_dii$p.adj <- stats::p.adjust(res_dii$P, method = "BH", n = n) 
```

There are no significant relationships after adjusting for multiple corrections in the full cohort. Prior to multiple testing correction, there is a significant finding with Vitamin C but that is not clearly supported by the literature.

Save for your records.

```{r save food linear regression table}
# Save table
write.csv(res_dii, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_DII.csv")
```

Double check by sex

```{r TMAO ~ dii - LCMS BD1 male and female - lm 356 ppl, results='hide'}
transformed_keep_tbl2 <- c("sex", 
#"sex_numeric",
"kcal_tnfs", 
"age", 
"cystatinc_bd1", 
"ethnicity",
"tmao", 
"tmao_log", 
"dii_alcohol",
"dii_b12",
"dii_b6",
"dii_betacarotene",
"dii_caffeine",
"dii_carb",
"dii_chol",
"dii_kcal",
"dii_totalfat",
"dii_fiber",
"dii_folicacid",
"dii_iron",
"dii_magnesium",
"dii_mufa",
"dii_niacin",
"dii_omega3",
"dii_omega6",
"dii_protein",
"dii_pufa",
"dii_riboflavin",
"dii_satfat",
"dii_selenium",
"dii_thiamin",
"dii_transfat",
"dii_vita",
"dii_vitc",
"dii_vitd",
"dii_vite",
"dii_zinc",
"dii_tea" ,
"dii_flav3ol",
"dii_flavone",
"dii_flavonol",
"dii_flavonon",
"dii_anthocy",     
"dii_isoflav",
"dii_total_score")

# Investigate by sex
#~~~~~~~~~~~~~
# FEMALES 
#~~~~~~~~~~~~~
# Make a smaller df from master (the old data)
full_sub_F <- full[, colnames(full) %in% transformed_keep_tbl2]
full_sub_F <- full_sub_F[full_sub_F$sex == "Female",]
dii_F <- subset(full_sub_F, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
n <- ncol(dii_F)
# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii_F[,x] + age + cystatinc_bd1 + kcal_tnfs, full_sub_F ))

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm356_lcms) <- colnames(dii_F)
res_dii_F <- as.data.frame(my_lm356_lcms)

# BH correction
res_dii_F$p.adj <- stats::p.adjust(res_dii_F$P, method = "BH", n = n)

#~~~~~~~~~~~~~
# MALES 
#~~~~~~~~~~~~~
# Make a smaller df from master (the old data)
full_sub_M <- full[, colnames(full) %in% transformed_keep_tbl2]
full_sub_M <- full_sub_M[full_sub_M$sex == "Male",]
dii_M <- subset(full_sub_M, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
n <- ncol(dii_M)
# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii_M[,x] + age + cystatinc_bd1 + kcal_tnfs, full_sub_M ))

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm356_lcms) <- colnames(dii_M)
res_dii_M <- as.data.frame(my_lm356_lcms)

# BH correction
res_dii_M$p.adj <- stats::p.adjust(res_dii_M$P, method = "BH", n = n) 
```

Still no significant results after BH correction. Prior to BH correction, there is a suggested relationship with folic acid and TMAO in males.

Make TMAO food intake by quantile table

```{r Food Intake by TMAO quantile}
# TMAO table by age bin - females
factorVars <- c("age_cat", "tmao_quantile")
vars <- c("pf_meat_totals", 
          "pf_curedmeat_totals",
          "pf_poult_totals",
          "pf_seafd_hi_totals",
          "pf_seafd_low_totals",
          "pf_eggs_totals",
          "pf_nutsds_totals",
          "pf_legumes_totals",
          "pf_total_totals",
          #"pf_soy_totals_Binary",
          "v_drkgr_totals",
          "v_redor_total_totals",
          #"v_redor_tomato_totals",
          #"v_redor_other_totals",
          "v_starchy_total_totals",
          #"v_starchy_potato_totals",
          #"v_starchy_other_totals",
          "v_other_totals",
          "v_total_totals",
          #"v_legumes_total",
          "g_whole_totals",
          "g_refined_totals",
          "g_total_totals",
          "d_total_totals",
          "d_milk_totals",
          "d_cheese_totals",
          "d_yogurt_totals",
          "hei_asa24_totalscore",
          "hei_asa24_totalveg",
          "hei_asa24_sfat",
          "choln_tnfs",
          "age",
          "ethnicity" ,
          "bmi_final",
          "tmao",
          "choline",
          "betaine",
          "carnitine")
table_food.fq <- CreateTableOne(vars = vars,
                                    strata = "tmao_quantile",
                                    data = full[full$sex == "Female",],
                                    factorVars = factorVars)
table_food.fq
